Initial estuarine response to the nutrient dense Piney Point wastewater discharge into Tampa Bay, Florida

### 2021-08-31 manu-draft.Rmd
separate pca and cor matrix for all variables, need to update text results

A PCA comparison of all water quality variables with seagrass and macroalgae frequency occurrence estimates shows significant assocations between groups of variables (Figure \@ref(fig:allpcacors)).  The first principal component axis explained 31\% of the variation among all variables, whereas the second and third axes explained 33\% of the remaining variation.  For the water quality varialbes, the first principal component axis had positive loadings for secchi depth and dissolved oyxgen and negatative assocations for total nitrogen, chl-a, and temperature.  For macroalgae, the first axis was positively associated with frequency occurrence of red and green macroalgae and negatively associated with cyanobacteria. Loading vectors for the seagrasses were minor, although turtle grass (*T. testudinum*) had a weak positive association with axis 1.  The second principal component axis had a strong postive assocation with total phosphorus and negative association with salinity. The third axis showed a strong positive association with manatee grass (*S. filiforme*) and negative association with shoal grass (*H. wrightii*).

Pairwise Spearman rank correlations for the same variables used in the PCA showed significant associations among several of the water quality, seagrass, and macroalgae variables.  Chlorophyll-a was positively correlated with total phosphorus, total nitrogen, temperature, and pH (all $p < 0.005$) and negatively correlated with secchi depth ($p < 0.005$).  Chlorophyll-a was also negatively correlated red and green macroalgae ($p < 0.05$) and turtle grass ($p < 0.005$).  Among the macroalgae, green and red groups were positively correlated ($p < 0.05$), whereas both were negatively correlated with cyanobacteria ($p < 0.005$ for red, $p < 0.05$ for green).  For water quality and seagrass, chl-a, total nitrogen, and total phosphorus was negatively correlated to turtle grass ($p < 0.005$ for all, except $p < 0.05$ for total nitrogen), whereas secchi was positively associated with shoal grass ($p < 0.05$).  For macroalgae and seagrass, red and cyanobacteria macroalgae were positively correlated with manatee grass ($p < 0.005$ for both), whereas cyanobacteria was negatively correlated with shoal grass ($p < 0.005$) and turtle grass ($p< 0.05$).  Among the seagrasses, manatee grass was negatively correlated with turtle grass and shoal grass ($p < 0.005$ for both).

@@

```{r allpcacors, fig.cap = 'Principal components analysis (PCA) and correlation matrix for water quality variables, macroalgae, and seagrasses by week for April to July 2021 in response to wastewater discharge from Piney Point.  Only areas 1 and 3 (Figure \\@ref(fig:map)a) are included where seagrass transects were surveyed. All variables were standardized to zero mean and unit variance prior to PCA. Variables with log-normal distribution were log-transformed prior to analysis.  Thick black lines in the correlation matrix separate water quality, macroalgae, and seagrass variables. Pearson correlations indicate the linear strength of assocation between pairs of variables. ** p < 0.005, * p < 0.05, blank is not significant at $\\alpha$ = 0.05.'}
knitr::include_graphics(here('figs/allpcacors.jpeg'))
```

>>>>

A PCA comparison of all water quality variables with seagrass and macroalgae frequency occurrence estimates shows significant assocations between groups of variables (Figure \@ref(fig:allpca)).  The first principal component axis explained 31\% of the variation among all variables, whereas the second and third axes explained 33\% of the remaining variation.  For the water quality varialbes, the first principal component axis had positive loadings for secchi depth and dissolved oyxgen and negatative assocations for total nitrogen, chl-a, and temperature.  For macroalgae, the first axis was positively associated with frequency occurrence of red and green macroalgae and negatively associated with cyanobacteria. Loading vectors for the seagrasses were minor, although turtle grass (*T. testudinum*) had a weak positive association with axis 1.  The second principal component axis had a strong postive assocation with total phosphorus and negative association with salinity. The third axis showed a strong positive association with manatee grass (*S. filiforme*) and negative association with shoal grass (*H. wrightii*).

Pairwise Spearman rank correlations for the same variables used in the PCA showed significant associations among several of the water quality, seagrass, and macroalgae variables (Figure \@ref(fig:allcor)).  Chlorophyll-a was positively correlated with total phosphorus, total nitrogen, temperature, and pH (all $p < 0.005$) and negatively correlated with secchi depth ($p < 0.005$).  Chlorophyll-a was also negatively correlated red and green macroalgae ($p < 0.05$) and turtle grass ($p < 0.005$).  Among the macroalgae, green and red groups were positively correlated ($p < 0.05$), whereas both were negatively correlated with cyanobacteria ($p < 0.005$ for red, $p < 0.05$ for green).  For water quality and seagrass, chl-a, total nitrogen, and total phosphorus was negatively correlated to turtle grass ($p < 0.005$ for all, except $p < 0.05$ for total nitrogen), whereas secchi was positively associated with shoal grass ($p < 0.05$).  For macroalgae and seagrass, red and cyanobacteria macroalgae were positively correlated with manatee grass ($p < 0.005$ for both), whereas cyanobacteria was negatively correlated with shoal grass ($p < 0.005$) and turtle grass ($p< 0.05$).  Among the seagrasses, manatee grass was negatively correlated with turtle grass and shoal grass ($p < 0.005$ for both).

@@

```{r allpca, fig.cap = 'Principal components analysis (PCA) for water quality variables, macroalgae, and seagrasses by week for April to July 2021 in response to wastewater discharge from Piney Point.  Only areas 1 and 3 (Figure \\@ref(fig:map)a) are included where seagrass transects were surveyed. All variables were standardized to zero mean and unit variance prior to PCA. Variables with log-normal distribution were log-transformed prior to analysis.'}
knitr::include_graphics(here('figs/allpca.jpeg'))
```

```{r allcor, fig.cap = 'Correlation matrix for water quality variables, macroalgae, and seagrasses by week for April to July 2021 in response to wastewater discharge from Piney Point.  Only areas 1 and 3 (Figure \\@ref(fig:map)a) are included where seagrass transects were surveyed. Variables with log-normal distribution were log-transformed prior to analysis.  Thick black lines separate water quality, macroalgae, and seagrass variables. Spearman correlations indicate the linear strength of assocation between pairs of variables. ** p < 0.005, * p < 0.05, blank is not significant at $\\alpha$ = 0.05.'}
knitr::include_graphics(here('figs/allcor.jpeg'))
```
